Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms

### 2018-09-05 paper/paper.tex
2019
Betterification

\def\year{2019}\relax
%File: formatting-instruction.tex
\documentclass[letterpaper]{article} %DO NOT CHANGE THIS
\usepackage{aaai18}  %Required
\usepackage{times}  %Required
\usepackage{helvet}  %Required
\usepackage{courier}  %Required
\usepackage{url}  %Required
\usepackage{graphicx}  %Required
\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:
  \pdfinfo{
/Title (Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms)
/Author (Anonymous and Anonymous and Anonymous)}
\setcounter{secnumdepth}{0}

@@

\begin{abstract}
    Modern subgraph isomorphism solvers use degree-based value-ordering heuristics to direct
    backtracking search. This kind of search makes a heavy commitment to the first branching choice,
    which is often incorrect. We evaluate discrepancy-based searches as an alternative, but find
    them too expensive to be effective in this situation, even on satisfiable instances. We then
    demonstrate that a random value-ordering heuristic with restarts beats conventional search with
    the best-known tailored heuristic. This motivates the introduction of a slightly-random
    heuristic with similar behaviour to the tailored heuristic: when combined with restarts and
    nogood recording, this new search strategy is over a hundred times more effective on satisfiable
    instances, and does not affect performance on unsatisfiable instances.  Finally, we show that
    this strategy also improves two maximum common induced subgraph algorithms.
\end{abstract}

@@

The subgraph isomorphism problem is to decide whether a copy of a small ``pattern'' graph occurs
inside a larger ``target'' graph. Although \NP-complete, by combining design techniques from
artificial intelligence with careful algorithm engineering, modern subgraph isomorphism solvers can
often produce exact solutions quickly even on graphs with thousands of vertices. The current single
strongest subgraph isomorphism solver uses ``highest degree first'' as a value-ordering heuristic to
direct a constraint programming style search
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:conf/ijcai/McCreeshPT16}. This
heuristic is much better than branching randomly, but is still far from perfect.  We could try to
introduce tiebreaking mechanisms and similar refinements to strengthen the heuristic. However, given
the extreme cost of making an incorrect branching choice early in search, we believe it is more
fruitful to investigate alternatives to simple backtracking (depth-first) search. This paper
evaluates two such approaches: discrepancy search (which performs poorly), and a new form of search
(which is extremely beneficial) based around restarts and nogood recording. To explore different
parts of the search space when restarting, we introduce a novel way of adapting a value-ordering
heuristic to add a very small amount of randomness. The effects of these changes are shown in the
final plot of \cref{figure:scatter-random}: we can make a state-of-the-art algorithm perform much
better on a large number of satisfiable instances, whilst performing worse only rarely on
satisfiable instances, and never on unsatisfiable instances.  Further, these properties are
\emph{not} shared by discrepancy searches (\cref{figure:scatter-dds}) or by random value-ordering
with restarts (the first two plots of \cref{figure:scatter-random}). Finally, we show that this
strategy is also effective in an optimisation setting, producing benefits in two maximum common
induced subgraph algorithms.

@@

Let $G$ be a graph. We denote its set of vertices by $\vertexset(G)$, and write $v \sim_G w$ to mean
that vertices $v$ and $w$ are adjacent; we allow a vertex to be adjacent to itself. The
\emph{neighbourhood} of a vertex, $\neighbourhood_G(v)$, is the set of vertices adjacent to $v$, and
the \emph{degree} of $v$, $\deg_G(v)$, is the cardinality of its neighbourhood. When the graph is clear from the context, we omit
the $G$ subscripts. As per \citet{DBLP:conf/aaai/HoffmannMR17}, $G^{n,\ell}$ denotes the graph with
vertex set $\vertexset(G)$, and with edges between vertices $v$ and $w$ if there are at least $n$
simple paths of length $\ell$ between vertices $v$ and $w$ in $G$.

The non-induced subgraph isomorphism problem is to find an injective mapping from the vertices of a
pattern graph $\mathcal{P}$ to a target graph $\mathcal{T}$, such that adjacent vertices in
$\mathcal{P}$ are mapped to adjacent vertices in $\mathcal{T}$ (including that vertices with loops
in $\mathcal{P}$ may only be mapped to vertices with loops in $\mathcal{T}$). We write $v \mapsto
v'$ to mean that pattern vertex $v$ is mapped to target $v'$ under such a mapping. Finally, we
define a function $\FuncSty{compatible}(v, w)$ to mean that mapping $v \mapsto w$ is feasible based
upon both the loop rule, and the neighbourhood degree sequence relation of
\citet{DBLP:journals/constraints/ZampelliDS10}.

@@

We now briefly outline the non-shaded parts of \cref{algorithm:sip}, which is how we will solve the
subgraph isomorphism problem. This algorithm is very closely based upon the $k{\downarrow}$
algorithm of \citet{DBLP:conf/aaai/HoffmannMR17} with $k = 0$, and we refer the reader to that paper
for full technical details; that algorithm, in turn, is a simplification of the Glasgow2 algorithm
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}.

@@

    \nl $\FuncSty{search}$ (Domains $D$, Decisions $B$) $\rightarrow$ Bool

@@

        \nl \ForEach{$D_w \in D - D_v$}{
            \nl $D_w \gets D_w - v'$ \siplabel{removev} \;

@@

For the variable-ordering heuristic on \siplineref{variableordering} of
\cref{algorithm:sip}, we use smallest domain first, tiebreaking on highest degree
\cite{DBLP:journals/ai/HaralickE80,DBLP:journals/jair/McCreeshPST18}. Variable ordering is not the
focus of this paper---as far as possible, we use this same dynamic ordering throughout the paper, so
as not to affect the performance of the algorithm on unsatisfiable instances.  For value ordering,
which occurs on \siplineref{valueordering} of \cref{algorithm:sip}, we will initially select
vertices from highest degree to lowest degree; we return to value ordering in the second half of
this paper. Note that the variable-ordering heuristic affects both unsatisfiable and satisfiable
instances, whilst the value-ordering heuristic is relevant only for satisfiable instances.

@@

Our experiments are performed on systems with dual Intel Xeon E5-2697A v4 CPUs and 512GBytes RAM,
running Ubuntu 17.04. We
implemented\footnote{DOI removed for anonymous review}
\cref{algorithm:sip} in C++ using the bit-parallel implementation of
\citet{DBLP:conf/aaai/HoffmannMR17} as a starting point, and compiled it using GCC 6.3.0. We use a
deterministic pseudo-random number generator for reproducibility.

@@

\noindent
Other studies use a random selection of 200 of each of the instances from the ``meshes'' and
``images'' families because some earlier solvers find many of these instances extremely hard. We
would like to have a larger number of satisfiable instances in our test set, and so we include all
pattern / target pairs. This gives a total of 14,621 instances (rather than the original 5,725). At
least \input{gen-n-sat.tex}\unskip\ of these instances are known to be satisfiable, and at least
\input{gen-n-unsat.tex}\unskip\ are unsatisfiable.

@@

We begin by verifying that our implementation of the basic \cref{algorithm:sip} is competitive. In
\cref{figure:others} we show the cumulative number of instances solved over time, using our
implementation (both with the degree heuristic, and with the modifications described in the
remainder of this paper), the Glasgow2 and Glasgow3 algorithms from which \cref{algorithm:sip} is derived
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}, the PathLAD variation of the LAD
algorithm \cite{DBLP:journals/ai/Solnon10,DBLP:conf/lion/KotthoffMS16}, and VF2
\cite{DBLP:journals/pami/CordellaFSV04}, in each case using the original implementation provided
by the algorithm's authors. For each algorithm, the $y$ value
gives the cumulative number of instances which (individually) can be solved in no more than $x$
milliseconds.  The vertical distance between two lines therefore shows how many more instances can
be solved by one solver than another, if every instance is run separately with the chosen $x$
timeout. The horizontal distance shows how many times longer the per-instance timeout would need to
be to allow the rightmost algorithm to succeed on $y$ out of the 14,621 instances (bearing in mind
that the two sets of $y$ instances could be different).

As expected, the performance of \cref{algorithm:sip} with the degree heuristic is very close to that
of Glasgow2, although we perform better at lower runtimes because we use simpler preprocessing at
the top of search. Our implementation also clearly beats PathLAD and VF2, except for very low
choices of timeout.

The dataset includes many instances which are extremely easy for a good solver, and so it can be
hard to see the differences between the stronger solvers at higher runtimes. This paper focusses
upon improving the performance on the remaining hard satisfiable instances, and so in the second
plot in \cref{figure:others} (and in subsequent cumulative plots) we show only satisfiable
instances, and use a reduced range on both axes.

@@

\Cref{figure:value-ordering-heuristics} shows the cumulative number of satisfiable instances
solved over time using different value-ordering heuristics:

@@

\item[Biased] We branch by selecting a vertex $v'$ from the chosen domain $D_v$ with
probability \[ p(v') = \frac{2^{\deg(v')}}{\sum_{w \in D_v}{2^{\deg(w)}}} \text{.} \] This is a new
heuristic, which continues to prefer vertices of high degree, but will give the same chance of
being selected next to two vertices of equal degree.  It also introduces an element of randomness,
which will be needed to make restarts have an effect. \end{description}

\noindent
The performance of the degree and biased heuristics are very similar: the
biased heuristic has slightly worse performance below a 400s timeout, and effectively equal
performance above it. This shows that we can introduce an element of randomness into the degree
value-ordering heuristic without adversely affecting its performance \emph{in aggregate}.
\Cref{figure:scatter-heuristics} gives a detailed look at these results. This scatter plot
shows that despite the aggregate performance being similar, on a case by case basis, the two
heuristics can make a large difference to the performance of satisfiable instances.

@@

A \emph{discrepancy} is where search goes against a value-ordering heuristic.  Discrepancy searches
\cite{DBLP:conf/ijcai/HarveyG95,DBLP:conf/aaai/Korf96,DBLP:conf/ijcai/Walsh97} are alternatives to
backtracking search that initially search disallowing all discrepancies, and then retry search
allowing an increasing number of discrepancies at each iteration until either a solution is found or
unsatisfiability is proven. These schemes assume that value-ordering heuristics are usually
reliable, and that most solutions can be found with only a small number of discrepancies; in such
cases, the heavy commitment to early branching choices made by conventional backtracking search can
be extremely costly.

\Cref{figure:scatter-dds} shows the effects of adding Walsh's \cite{DBLP:conf/ijcai/Walsh97}
depth-bounded discrepancy search (DDS) to \cref{algorithm:sip}. Each point represents the solving
time for one instance---to avoid noise for easier instances, we measure only time spent during
search, and exclude time spent in preprocessing and initialisation.  Points below the $x-y$ diagonal
are speedups, whilst points on the top and right axes represent instances which timed out after one
thousand seconds with one algorithm, but not the other. For satisfiable instances, the different
point styles show the different families, whilst all unsatisfiable instances are shown as dark dots.

The points well below the diagonal line and along the right-hand axis show that DDS can sometimes be
extremely beneficial on satisfiable instances.  However, on both unsatisfiable and most satisfiable
instances, the overheads can be extremely large, and DDS is much worse in aggregate (even when only
considering satisfiable instances).

These large overheads are to be expected: discrepancy searches are aimed primarily at getting better
feasible solutions in optimisation problems which are too large for a proof of optimality to be a
realistic prospect, and they are not well-suited for unsatisfiable decision problems. However, the
extremely large gains on some satisfiable instances confirm our suspicions that we should find an
alternative to heavy-commitment backtracking search.

@@

Another alternative to plain backtracking search is provided by \emph{restarts}. The general idea is
to perform a certain amount of search, and then if no solution has been found (and unsatisfiability
has not been proven), to abandon search and restart from the beginning. Obviously, such an approach
can only be beneficial if something changes after restarting---in a constraint programming setting,
this is usually the variable-ordering heuristic
\cite{DBLP:journals/jsat/LecoutreSTV07,DBLP:conf/cp/GayHLS15,DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}.
In this paper, we instead rely upon randomness in the \emph{value}-ordering heuristic, and continue
to use smallest domain first with static tiebreaking for variable-ordering.\footnote{It may be
possible to further improve \cref{algorithm:sip} by also introducing randomness or some form of
learning into its variable-ordering heuristic. However, simultaneously introducing a second change
would considerably complicate the empirical analysis.} Using restarts on value-ordering heuristics
appears to be uncommon---although \citet{DBLP:conf/flairs/RazgonOP07} look at learning
value-ordering heuristics from restarts, \citet{DBLP:conf/cp/ChuSS09} use a similar
scheme in the context of parallel search, and an early approach by \citet{DBLP:conf/aaai/GomesSK98} does so in an optimisation context.

@@

\Cref{figure:restarts} also shows that the biased heuristic together with restarts is better
still---that is, if we are introducing restarts, then it is better to add a small amount of
randomness to a tailored deterministic heuristic than it is to throw away the heuristic altogether.
Indeed, the original algorithm can solve 1,983 satisfiable instances by 993.0 seconds, whilst the
biased and random restarting algorithms require only 9.2 seconds and 33.8 seconds respectively to
solve the same number.

In the more detailed view in the final plot of \cref{figure:scatter-random}, comparing the
basic algorithm with the degree heuristic to the degree-biased algorithm with restarts, all of the
unsatisfiable instances are very close to the $x-y$ diagonal, showing that their performance is
nearly unchanged. On the other hand, there are large numbers of satisfiable instances well below the
diagonal line, indicating large speedups.  Better yet, there are only a handful of satisfiable
instances that are more than a factor of ten times worse. In other words, as well as improving
performance, we have made up most of the consistency we lost by introducing randomness.

@@

The effectiveness of this approach raises a number of interesting questions and avenues for possible
future research. For example, will this same technique work for other problems where we have a good
value-ordering heuristic? Are there further benefits to be had from also modifying variable-ordering
heuristics? Are there theoretical rather than just empirical reasons for using an exponential bias
scheme? And how can such a search process be parallelised?

>>>>

\def\year{2019}\relax
%File: formatting-instruction.tex
\documentclass[letterpaper]{article} %DO NOT CHANGE THIS
\usepackage{aaai19}  %Required
\usepackage{times}  %Required
\usepackage{helvet}  %Required
\usepackage{courier}  %Required
\usepackage{url}  %Required
\usepackage{graphicx}  %Required
\frenchspacing  %Required
\setlength{\pdfpagewidth}{8.5in}  %Required
\setlength{\pdfpageheight}{11in}  %Required
%PDF Info Is Required:
  \pdfinfo{
/Title (Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms)
/Author (Anonymous and Anonymous and Anonymous)}
\setcounter{secnumdepth}{0}

@@

\begin{abstract}
    Modern subgraph isomorphism solvers use degree-based value-ordering heuristics to direct
    backtracking search. This kind of search makes a heavy commitment to the first branching choice,
    which is often incorrect. We evaluate discrepancy-based searches as an alternative, but find
    them too expensive to be effective in this situation, even on satisfiable instances. We then
    demonstrate that a random value-ordering heuristic with restarts beats conventional search with
    the best-known tailored heuristic. This motivates the introduction of a novel slightly-random
    heuristic with similar behaviour to the tailored heuristic: when combined with restarts and
    nogood recording, this new search strategy is over a hundred times more effective on satisfiable
    instances, and does not affect performance on unsatisfiable instances.  Finally, we show that
    this strategy also improves two maximum common induced subgraph algorithms.
\end{abstract}

@@

The subgraph isomorphism problem is to decide whether a copy of a small ``pattern'' graph occurs
inside a larger ``target'' graph. Although \NP-complete, by combining design techniques from
artificial intelligence with careful algorithm engineering, modern subgraph isomorphism solvers can
often produce exact solutions quickly even on graphs with thousands of vertices. The current single
strongest subgraph isomorphism solver uses ``highest degree first'' as a value-ordering heuristic to
direct a constraint programming style search
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:journals/jair/McCreeshPST18}. This
heuristic is much better than branching randomly, but is still far from perfect. We therefore
investigate different techniques for offsetting mistakes in value-ordering decisions. We first try
discrepancy searches, which turn out to be much too expensive to be a practical solution. We then
look at using random value-ordering, restarts, and nogood recording, which together give better
aggregate performance at the expense of much worse behaviour on many instances. Next, we introduce a
new way of adapting the existing value-ordering heuristic to introduce just a small amount of
randomness.  This lets us make a state-of-the-art subgraph algorithm perform much better
on a large number of satisfiable instances, whilst performing worse only rarely on satisfiable
instances, and never on unsatisfiable instances. Finally, we show that this strategy is also
effective in an optimisation setting, producing benefits in two maximum common induced subgraph
algorithms. This combination of techniques gives us, for the first time, a practical alternative to
backtracking search for constraint programming style algorithms where we have a strong
value-ordering heuristic and where we care both about satisfiable and unsatisfiable instances.

@@

Let $G$ be a graph. We denote its set of vertices by $\vertexset(G)$, and write $v \sim_G w$ to mean
that vertices $v$ and $w$ are adjacent; we allow a vertex to be adjacent to itself (a \emph{loop}). The
\emph{neighbourhood} of a vertex, $\neighbourhood_G(v)$, is the set of vertices adjacent to $v$, and
the \emph{degree} of $v$, $\deg_G(v)$, is the cardinality of its neighbourhood. When the graph is clear from the context, we omit
the $G$ subscripts. As per \citet{DBLP:conf/aaai/HoffmannMR17}, $G^{n,\ell}$ denotes the graph with
vertex set $\vertexset(G)$, and with edges between vertices $v$ and $w$ if there are at least $n$
simple paths of length $\ell$ between vertices $v$ and $w$ in $G$.

The \emph{non-induced subgraph isomorphism problem} is to find an injective mapping from the
vertices of a pattern graph $\mathcal{P}$ to a target graph $\mathcal{T}$, such that adjacent
vertices in $\mathcal{P}$ are mapped to adjacent vertices in $\mathcal{T}$ (including that vertices
with loops in $\mathcal{P}$ may only be mapped to vertices with loops in $\mathcal{T}$). The
\emph{induced} problem additionally requires that non-adjacent vertices are mapped to non-adjacent
vertices. We write $v \mapsto v'$ to mean that pattern vertex $v$ is mapped to target $v'$ under
either kind of mapping. Finally, we define a function $\FuncSty{compatible}(v, w)$ to mean that
mapping $v \mapsto w$ is feasible based upon both the loop rule, and the neighbourhood degree
sequence relation of \citet{DBLP:journals/constraints/ZampelliDS10}.

@@

We now briefly outline the non-shaded parts of \cref{algorithm:sip}, which is how we will solve the
subgraph isomorphism problem. This algorithm is very closely based upon the $k{\downarrow}$
algorithm of \citet{DBLP:conf/aaai/HoffmannMR17} with $k = 0$, and we refer the reader to that paper
for full technical details; that algorithm, in turn, is a simplification of the Glasgow2 algorithm
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}. We show the non-induced problem; the
induced problem may be solved by including the loop-complements of $\mathcal{P}$ and $\mathcal{T}$
in $L$ on \siplineref{supplemental}.

@@

    \nl $\FuncSty{search}$ (Domains $D$, Trail $B$) $\rightarrow$ Bool

@@

        \nl \ForEach{$D_w \in D \setminus \{ D_v$ \}}{
            \nl $D_w \gets D_w \setminus \{ v' \}$ \siplabel{removev} \;

@@

For the variable-ordering heuristic on \siplineref{variableordering} of
\cref{algorithm:sip}, we use smallest domain first, tiebreaking on highest degree
\cite{DBLP:journals/ai/HaralickE80,DBLP:journals/jair/McCreeshPST18}. Variable ordering is not the
focus of this paper---as far as possible, we use this same dynamic ordering throughout the paper, so
as not to affect the performance of the algorithm on unsatisfiable instances.  For value ordering,
which occurs on \siplineref{valueordering} of \cref{algorithm:sip}, we will initially select
vertices from highest degree to lowest degree; value ordering is the major theme of the remainder of
this paper. Note that the variable-ordering heuristic affects both unsatisfiable and satisfiable
instances, whilst the value-ordering heuristic is relevant only for satisfiable instances.

@@

Our experiments are performed on systems with dual Intel Xeon E5-2697A v4 CPUs and 512GBytes RAM,
running Ubuntu 17.04. We
implemented\footnote{DOI removed for anonymous review}
\cref{algorithm:sip} in C++ using the bit-parallel implementation of
\citet{DBLP:conf/aaai/HoffmannMR17} as a starting point, and compiled it using GCC 7.2.0. We use a
deterministic pseudo-random number generator for reproducibility.

@@

\noindent
Other studies use a random selection of 200 of each of the instances from the ``meshes'' and
``images'' families because some earlier solvers find many of these instances extremely hard. We
would like to have a larger number of satisfiable instances in our test set, and so we include all
pattern / target pairs. This gives a total of 14,621 instances (rather than the original 5,725). At
least \input{gen-n-sat.tex}\unskip\ of these instances are known to be satisfiable for the
non-induced problem, and at least \input{gen-n-unsat.tex}\unskip\ are unsatisfiable.

@@

We begin by verifying that our implementation of the basic \cref{algorithm:sip} is competitive. In
the first and third plots \cref{figure:others} we show the cumulative number of instances solved
over time for the non-induced and induced problems respectively. We compare our implementation (both
with the degree heuristic, and with the modifications described in the remainder of this paper), the
Glasgow2 and Glasgow3 algorithms from which \cref{algorithm:sip} is derived
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}, the PathLAD variation of the LAD
algorithm \cite{DBLP:journals/ai/Solnon10,DBLP:conf/lion/KotthoffMS16}, VF2
\cite{DBLP:journals/pami/CordellaFSV04}, RI \cite{DBLP:journals/bmcbi/BonniciGPSF13}, and VF3
\cite{DBLP:conf/gbrpr/CarlettiFSV17} (which only supports the induced problem), in each case using
the original implementation provided by the algorithm's authors. For each algorithm, the $y$ value
gives the cumulative number of instances which (individually) can be solved in no more than $x$
milliseconds.  The vertical distance between two lines therefore shows how many more instances can
be solved by one solver than another, if every instance is run separately with the chosen $x$
timeout. The horizontal distance shows how many times longer the per-instance timeout would need to
be to allow the rightmost algorithm to succeed on $y$ out of the 14,621 instances (bearing in mind
that the two sets of $y$ instances could be different).

As expected, in both cases the performance of \cref{algorithm:sip} with the degree heuristic is very
close to that of Glasgow2, although we perform better at lower runtimes because we use simpler
preprocessing at the top of search. Our implementation also clearly beats PathLAD, VF2, VF3 and RI,
except for very low choices of timeout.

The dataset includes many instances which are extremely easy for a good solver, and so it can be
hard to see the differences between the stronger solvers at higher runtimes. This paper focusses
upon improving the performance on the remaining hard satisfiable instances, and so in the second and
fourth plots in \cref{figure:others} (and in subsequent cumulative plots) we show only satisfiable
instances, and use a reduced range on both axes.

@@

\Cref{figure:value-ordering-heuristics} shows the cumulative number of satisfiable instances
solved over time using different value-ordering heuristics. For the remainder of this section, we
show only the non-induced problem; results with the induced variant are similar.

@@

\item[Biased] We branch by selecting a vertex $v'$ from the chosen domain $D_v$ with
probability \[ p(v') = \frac{2^{\deg(v')}}{\sum_{w \in D_v}{2^{\deg(w)}}} \text{.} \]\end{description}

\noindent
The biased heuristic is new. It continues to prefer vertices of high degree, but will give the same
chance of being selected next to two vertices of equal degree.  It also introduces an element of
randomness, which will be needed to make restarts have an effect.  The performance of the degree and
biased heuristics are very similar: the biased heuristic has slightly worse performance below a 400s
timeout, and effectively equal performance above it. This shows that we can introduce an element of
randomness into the degree value-ordering heuristic without adversely affecting its performance
\emph{in aggregate} (and more generally, this scheme could be used for any value-ordering
heuristic which puts choices in some kind of rank with ties.)

\Cref{figure:scatter-heuristics} gives a detailed look at these results. This scatter plot shows
that despite the aggregate performance being similar, on a case by case basis, the two heuristics
can make a large difference to the performance of satisfiable instances.

@@

A \emph{discrepancy} is where search goes against a value-ordering heuristic.  Discrepancy searches
\cite{DBLP:conf/ijcai/HarveyG95,DBLP:conf/aaai/Korf96,DBLP:conf/ijcai/Walsh97,DBLP:conf/cpaior/KarouiHLN07}
are alternatives to backtracking search that initially search disallowing all discrepancies, and
then retry search allowing an increasing number of discrepancies at each iteration until either a
solution is found or unsatisfiability is proven. These schemes assume that value-ordering heuristics
are usually reliable, and that most solutions can be found with only a small number of
discrepancies; in such cases, the heavy commitment to early branching choices made by conventional
backtracking search can be extremely costly.

\Cref{figure:scatter-dds} shows the effects of adding Walsh's \cite{DBLP:conf/ijcai/Walsh97}
depth-bounded discrepancy search (DDS) to \cref{algorithm:sip} (results with other discrepancy
search variants are similar). Each point represents the solving
time for one instance---to avoid noise for easier instances, we measure only time spent during
search, and exclude time spent in preprocessing and initialisation.  Points below the $x-y$ diagonal
are speedups, whilst points on the top and right axes represent instances which timed out after one
thousand seconds with one algorithm, but not the other. For satisfiable instances, the different
point styles show the different families, whilst all unsatisfiable instances are shown as dark dots.

The points well below the diagonal line and along the right-hand axis show that DDS can sometimes be
extremely beneficial on satisfiable instances.  However, on both unsatisfiable and most satisfiable
instances, the overheads can be extremely large, and DDS is much worse in aggregate and is not a
viable approach (even when only considering satisfiable instances).  These large overheads are to be
expected: discrepancy searches are aimed primarily at getting better feasible solutions in
optimisation problems which are too large for a proof of optimality to be a realistic prospect, and
they are not well-suited for unsatisfiable decision problems. Despite this, the extremely large gains on
some satisfiable instances confirm our suspicions that we should find an alternative to
heavy-commitment backtracking search.

@@

Another alternative to plain backtracking search is provided by \emph{restarts}. The general idea is
to perform a certain amount of search, and then if no solution has been found (and unsatisfiability
has not been proven), to abandon search and restart from the beginning. Obviously, such an approach
can only be beneficial if something changes after restarting---in a constraint programming setting,
this is usually the variable-ordering heuristic
\cite{DBLP:journals/jsat/LecoutreSTV07,DBLP:conf/cp/GayHLS15,DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}.
In this paper, we instead rely upon randomness in a special new \emph{value}-ordering heuristic, and continue
to use smallest domain first with static tiebreaking for variable-ordering.\footnote{It may be
possible to further improve \cref{algorithm:sip} by also introducing randomness or some form of
learning into its variable-ordering heuristic. However, simultaneously introducing a second change
would considerably complicate the empirical analysis.} Using restarts on value-ordering heuristics
appears to be uncommon---although \citet{DBLP:conf/flairs/RazgonOP07} look at learning
value-ordering heuristics from restarts, \citet{DBLP:conf/cp/ChuSS09} use a similar
scheme in the context of parallel search, and an early approach by \citet{DBLP:conf/aaai/GomesSK98} does so in an optimisation context.

@@

\Cref{figure:restarts} also shows that the biased heuristic together with restarts is better
still---that is, if we are introducing restarts, then it is better to add a small amount of
randomness to a tailored heuristic than it is to throw away the heuristic altogether.
Indeed, the original algorithm can solve 1,983 satisfiable instances by 993.0 seconds, whilst the
biased and random restarting algorithms require only 9.2 seconds and 33.8 seconds respectively to
solve the same number.

In the more detailed view in the final plot of \cref{figure:scatter-random}, comparing the
basic algorithm with the degree heuristic to the degree-biased algorithm with restarts, all of the
unsatisfiable instances are very close to the $x-y$ diagonal, showing that their performance is
nearly unchanged. On the other hand, there are large numbers of satisfiable instances well below the
diagonal line, indicating large speedups.  Better yet, there are only a handful of satisfiable
instances that are more than a factor of ten times worse. In other words, as well as improving
performance, we have made up most of the consistency we lost by introducing randomness. This is a
promising result: for the first time, we have introduced a viable alternative to backtracking search
in a setting where we both have a strong \emph{value}-ordering heuristic, \emph{and} care about both
satisfiable and unsatisfiable instances.

@@

The effectiveness of this approach raises a number of interesting questions and avenues for possible
future research. For example, will this same technique work for other problems where we have a good
value-ordering heuristic? Are there further benefits to be had from also modifying variable-ordering
heuristics? And how can such a search process be parallelised?
